### 2024-12-03 00_T-reX_journal-upload-folder/T-reX_latex-files/T-reX_manuscript-review_changes.tex
add revision comparisons

% If the frontmatter runs over more than one page
% use the longmktitle option.
%\documentclass[a4paper,fleqn,longmktitle]{cas-dc}

@@

%\usepackage[numbers]{natbib}
\usepackage[authoryear]{natbib}
% \usepackage[authoryear,longnamesfirst]{natbib}
%\usepackage{booktabs}
%\usepackage[tt=false]{libertine}
%\usepackage{xurl}

%%%
\begin{document}
	\let\WriteBookmarks\relax
	\def\floatpagepagefraction{1}
	\def\textpagefraction{.001}

	% Short title
	\shorttitle{T-reX: Quantifying waste and material footprints in LCA}

	% Short author
	\shortauthors{McDowall, S2. et~al.}

	% Main title of the paper
	\title [mode = title]{T-reX: Quantifying waste and material footprints in current and future Life Cycle Assessment (LCA) databases}

	% First author
	%
	% Options: Use if required
	% eg: \author[1,3]{Author Name}[type=editor,
	%       style=chinese,
	%       auid=000,
	%       bioid=1,
	%       prefix=Sir,
	%       orcid=0000-0000-0000-0000,
	%       facebook=<facebook id>,
	%       twitter=<twitter id>,
	%       linkedin=<linkedin id>,
	%       gplus=<gplus id>]

@@

	% Corresponding author indication
	\cormark[1]
	% Email id of the first author
	\ead{s.c.mcdowall@cml.leidenuniv.nl}
	% URL of the first author
	\ead[url]{scmcd.ch}
	% Credit authorship
	\credit{Methodology, Software, Validation, Formal analysis, Investigation, Data curation, Writing: original draft, Writing: review \& editing, Visualization}

@@

	% Corresponding author text
	\cortext[1]{Corresponding author}

	% Here goes the abstract
	\begin{abstract}
		The quintessential principle of the circular economy is to minimise material consumption and waste generation. Thus, identifying and quantifying waste and material flows is critical.

		Life Cycle Assessment (LCA) is a powerful quantitative method for this, especially given its capacity to expose the details of an activity's entire life cycle, often guiding the implementation of circular principles to where they could be most effective.

		Pursuant to these goals the authors have developed `T-reX', an analytical tool for LCA, written in Python. T-reX extends on the existing computational ecosystem centered on \texttt{Brightway} and \texttt{Premise} and can facilitate the quantification of user-defined supply chain demands for activities in current and prospective scenarios. T-reX streamlines database manipulation for LCA practitioners, integrating methods to aggregate and analyse demand inventories that could expose hotspots of hidden risk.

		With a simple case study using lithium-ion batteries, we demonstrate some of T-reX's potential by detailing and exploring aspects of their waste and material footprints. Since such footprints are often linked to negative externalities, T-reX could support sustainable decision-making and contribute to the development of the circular economy by promoting the understanding of our material consumption and waste generation.
	\end{abstract}

	% Use if graphical abstract is present
	\begin{graphicalabstract}
		\includegraphics[width=\columnwidth]{grabs_T-reX_graphical-abstract.pdf}
		%\includegraphics[width=\columnwidth]{grabs_T-reX_graphical-abstract_MID-RES.png}
	\end{graphicalabstract}

	% Highlights
	\begin{highlights}
		\item T-reX, a new open-source program for waste \& material footprints in Life Cycle Assessment (LCA)
		\item Assesses supply risks by calculating demand inventories for critical materials
		\item Simplifies quantification of user-specified waste and material footprint categories
		\item Rapidly identifies hotspots of waste generation and material demand in supply	chains
		\item Presents a case study of five lithium-ion batteries to demonstrate T-reX's potential
	\end{highlights}

	% Keywords
	% Each keyword is seperated by \sep

	\begin{keywords}
		circular economy, waste footprint, material footprint, life cycle assessment, critical raw materials, supply chain
	\end{keywords}

@@

	% Main text
	\begin{table*}[!htbp]
		\centering
		\caption{List of abbreviations}\label{tab:abbreviations}
		\begin{tabular}{ll}
			\toprule
			CCS	              & Carbon Capture and Storage                                          \\
			CF                & Carbon Footprint                                                    \\
			EF                & Ecological Footprint                                                \\
			CML               & Centrum voor Milieuwetenschappen (Centre for Environmental Science) \\
			CRM               & Critical Raw Material                                               \\
			CPC               & Cooperative Patent Classification                                   \\
			CSI               & Crustal Scarcity Indicator                                          \\
			CSP               & Crustal Scarcity Potential                                          \\
			EDIP              & Environmental Design of Industrial Products                         \\
			EEIOA             & Environmentally Extended Input-Output Analysis                      \\
			EOL               & End-of-Life                                                         \\
			FOEN              & Swiss Federal Office for the Environment                            \\
			IAM               & Integrated Assessment Model                                         \\
			IE                & Industrial Ecology                                                  \\
			LCA               & Life Cycle Assessment                                               \\
			LCI               & Life Cycle Inventory                                                \\
			LCIA              & Life Cycle Impact Assessment                                        \\
			LFP               & Lithium Iron Phosphate                                              \\
			Li-ion            & Lithium-ion                                                         \\
			LiMn\(_2\)O\(_4\) & Lithium Manganese Oxide                                             \\
			MF                & Material Footprint                                                  \\
			MFA               & Material Flow Analysis                                              \\
			NCA               & Nickel Cobalt Aluminum                                              \\
			NMC               & Nickel Manganese Cobalt                                             \\
			PWF               & Product Waste Footprint                                             \\
			RCP               & Representative Concentration Pathway                                \\
			ReCiPe            & A standard LCIA method set                                          \\
			SDG               & Sustainable Development Goals                                       \\
			SSPs              & Shared Socioeconomic Pathways                                       \\
			T-reX             & The Tool for analysing re-X in LCA                                  \\
			UBP               & Umweltbelastungspünkte (Environmental Impact Points)                \\
			UN                & United Nations                                                      \\
			WF                & Waste Footprint                                                     \\
			\bottomrule
		\end{tabular}
	\end{table*}

	\section{Introduction}\label{sec:introduction}  
	\subsection{Introduction to T-reX}\label{sec:intro-trex}

@@

	\subsection{Background and need}\label{sec:intro-background}

@@

@@

	\begin{figure*}[]
		\centering
		\includegraphics[width=1.8\columnwidth]{T-reX_method.pdf}
		\caption{The workflow of T-reX presented in a generalised (a) and computational (b) format. Application of T-reX to one or more LCA databases creates a \texttt{brightway} project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain.}\label{fig:methods-flowchart}
	\end{figure*}

@@

	To provide an example, we investigated five types of Li-ion batteries in the
	\texttt{ecoinvent} 3.9.1 `cutoff' LCA database~\citep{ecoinvent2016version3},
	each represented by their unamended global market inventories. The system
	boundaries. These following markets were used for the example calculations and
	unaltered by the authors:

	\begin{itemize}
		\item Li-ion, LFP, rechargeable, prismatic
		\item Li-ion, LiMn\(_2\)O\(_4\), rechargeable, prismatic
		\item Li-ion, NMC111, rechargeable, prismatic
		\item Li-ion, NMC811, rechargeable, prismatic
		\item Li-ion, NCA, rechargeable, prismatic
	\end{itemize}

@@

	\begin{itemize}
		\item ReCiPe 2016 v1.03, midpoint (I)~\citep{huijbregts2016recipe}
		\item Environmental Footprint (EF) v3.1 (no long-term emissions)~\citep{eu2023ef}
		\item Spatial differentiation in life cycle impact assessment---the EDIP 2003 no long term emissions method~\citep{hauschild2003edip}
		\item The crustal scarcity indicator method---CSI 2020~\citep{arvidsson2020csi}
	\end{itemize}

@@

	\subsubsection{LCA calculations}
	For each combination of activity, method, and database, a total summation was
	calculated along with details of the top contributing processes. Additionally,
	for the T-reX methods, a contribution analysis was performed with the
	\texttt{bwa.compare\_activities\_by\_grouped\_leaves} function from the
	\texttt{brightway2\_analyzer} package~\citep{mutel2016brightway2analyzer}, a
	component of the \texttt{brightway} ecosystem. This function performs one
	subjective method of graph traversal on the impact matrix of the LCA object to
	a specified cutoff. The groups of the resulting `leaves' are collected by their
	Cooperative Patent Classification (CPC) codes [giving a representation of
	industry sector involvement]. T-reX provides, essentially, a tabulation of
	internal exchanges and an insight into the products and sectors in the supply
	chain of the activity that carry the most substantial responsibility for the
	final waste generation or material demand footprints.

@@

	An example of the output from the application of T-reX has been included in the
	supplementary material~\autoref{sec:supplementary-5}. The
	manipulated \texttt{ecoinvent} databases (which are the main product of T-reX)
	can be recreated by following the instructions in the package
	documentation~\citep{mcdowall2023T-reXdocs}.

@@

	As described in~\autoref{sec:method-casestudy}, this case study calculated the
	waste and material footprints (with a variety of other indicators) for the
	unaltered background inventories of five Li-ion batteries with the functional
	unit being 1~kg of the battery at the global market. The purpose of this simple
	case study was to test, verify, and demonstrate the functionality and
	limitations of T-reX. This section includes some highlights of the results,
	with full results included in the supplementary material
	(\autoref{sec:supplementary-7} and \autoref{sec:supplementary-8}). Because the `pseudo-LCIA' methods created by
	T-reX are integrated into the user's \texttt{brightway} project as if they were
	standard LCIA methods, the footprint calculations can be performed in the
	customary way. In the supplementary material, there are screenshots of selected
	results obtained using the \texttt{ActivityBrowser} software, including
	contribution analysis and a Sankey diagram that dis-aggregates the final
	footprint result over the activities in the supply chain.

	\subsubsection{Sankey visualisation of flow inventory footprints}\label{sec:results-case_study-sankey}

	\autoref{fig:sankey_waste} presents a heavily abridged Sankey diagram depicting the supply chain flows of `total solid waste' attributed to the NMC 811 Li-ion battery at the global market in the \texttt{ecoinvent} database version 3.9.1. As the supply chains are expansive, this visualisation has been greatly simplified to be legible for print, multiple versions are included in the supplementary material in~\autoref{sec:supplementary-8}. As is evident in the figure, the largest contributors to the waste footprint are the extraction and processing of the raw materials, with the largest flows coming from the extraction of cobalt, lithium, and nickel. Due to the nature of LCA, the waste flows do not represent the `real' physical flows directly, but rather an abstraction thereof based on, in this case, economic allocation---the market price of the functional unit produced by the activity~\citep{guinee2004economicallocation}. If an unallocated database were available, the waste flows could be calculated with T-reX in the same way and would be more directly related to the physical flows of the activities.

	\begin{figure}[!htbp]
		\centering
		\includegraphics[width=0.8\columnwidth]{T-reX_NMC811_WasteTotalSolid.pdf}
		\caption{A greatly simplified Sankey diagram from \texttt{ActivityBrowser} showing the total solid waste footprint flows for the
			NMC 811 battery supply chain at the global market in the \texttt{ecoinvent} database version 3.9.1}\label{fig:sankey_waste}
	\end{figure}

@@

	%\label{sec:results-case_study-total_footprints}

	%! total
	\autoref{fig:waste_totalANDcarbon} (a) shows the `total solid waste' inventory footprint for the five Li-ion batteries in the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model. The NMC811 battery has the largest footprint, producing over 50~kg of waste per kilogram of battery produced. The  LiMn\(_2\)O\(_4\) battery has the smallest footprint, producing less than 4~kg of waste per kilogram of battery. In each case, there was only a slight downward trend in the waste footprints between 2020 and 2100. This is mostly in the period between 2020 and 2040 and could be attributable to the relatively rapid decrease in fossil fuel use that is factored into the models over this time (since many produce relatively large amounts of waste during extraction and combustion). For the total waste generated by these batteries, there was very little difference observed between the baseline and PkBudg500 RCPs.

	%! CO2
	The inclusion of carbon capture and storage (CCS) in the prospective databases
	is apparent in~\autoref{fig:waste_totalANDcarbon} (b), with the rapid increase
	in the production of carbon dioxide `waste' over the period from 2020--2040
	that is not seen in the baseline scenario. This result highlights the fact that
	the (often) downward trends in global warming impacts calculated with
	prospective databases using standard LCIA methods are dependent on the
	assumptions made about the introduction of CCS technology. The actual
	deployment of these technologies was only around 37~Mt CO\(_2\)/yr as of
	2023~\citep{dziejarski2023ccs}, far short of the levels projected in many of
	the RCP scenarios~\citep{sacchi2023premisedocs}. The authors consider that the
	over-representation of CCS and the under-representation of waste processing
	technologies in the prospective databases is one significant limitation to the
	validity of prospective LCA databases. Significant work is needed to improve
	the representation of these technologies in the databases to ensure that the
	results are more realistic and useful for decision-making.
	%!!!!

	\begin{figure*}[!htbp]
		\centering
		\includegraphics[width=1.8\columnwidth]{T-reX-wastefootprint-totalANDcarbon.pdf}
		\caption{T-reX calculated supply chain inventories for the five Li-ion batteries in the case study, showing their `footprints' for: (a) total solid waste, and (b) carbon dioxide waste (produced from carbon capture and storage (CCS) only). The footprints were modelled with prospective LCA databases from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:waste_totalANDcarbon}
	\end{figure*}

	\subsubsection{Contribution of `top-processes' in the supply chain}\label{sec:results-case_study-topprocesses}

@@

	\begin{figure}[!htbp]
		\centering
		\includegraphics[width=0.8\columnwidth]{T-reX-wastefootprint-processcontributions.pdf}
		\caption{Contribution by the `top-processes' to graphite demand inventory of the LiMn\(_2\)O\(_4\) battery from 2020 to 2100 under the SSP2 scenario using the baseline RCPs of the REMIND model}\label{fig:top_contribution}
	\end{figure}

@@

	\begin{figure}
		\centering
		\includegraphics[width=0.8\columnwidth]{T-reX-wastefootprint-sectorcontributions.pdf}
		\caption{Contribution of industrial sectors to the natural gas footprint of the LFP battery from 2020 to 2100 under the SSP2 scenario using the baseline RCP of the REMIND model}\label{fig:cpc_contribution}
	\end{figure}

@@

	A comparison of the results for `Coal (black)' demand using the T-reX method
	with the LCIA method `EDIP 2003 – coal no LT' is shown
	in~\autoref{fig:comparison_methods}. In this case, both the trends and the
	magnitude of the scores were very similar, both demonstrating a general
	decrease in the use of coking coal in the battery's footprints. Such
	comparability was also observed for other fossil-fuel-related demands (e.g.,
	natural gas and petroleum), and to a lesser extent, for some of the metal
	demand methods (e.g., zinc and cobalt). Correlation between standard LCIA
	methods and T-reX methods is not generally to be expected, however, due to
	fundamental differences in the way that the methods are constructed. In
	standard LCIA methods, impact scores are derived exclusively from the magnitude
	of the exchanges between the biosphere and the technosphere, that is,
	extraction and emission. In the T-reX methods, the footprint scores are based
	on an accounting of either waste generated or material demand, both of which
	are technosphere-technosphere exchanges in terms of LCA modelling. For the
	material demand methods especially, this distinction is critical. For example,
	the application of the EDIP 2003 or the CSP methods for a given metal will
	provide a score that is proportional to the amount extracted by mining, whereas
	the T-reX method provides an aggregation of the exchanges with the market for
	that metal. The T-reX method, therefore, considers cases of co-production,
	recycling, and substitution, providing a picture of the supply chain pressures
	that are not captured by the standard LCIA methods. This makes the T-reX
	methods more sensitive to the modeling choices (e.g., economic allocation) that
	are generally embedded in LCA databases.

	\begin{figure*}[!htbp]
		\centering
		\includegraphics[width=1.8\columnwidth]{T-reX-coalANDedip.pdf}
		\caption{Comparison of the (a) `T-reX – Coal (black)' demand footprint with the (b) conventional LCIA method `EDIP 2003 – coal no longterm'. In the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model for all five batteries.}\label{fig:comparison_methods}
	\end{figure*}

@@

	% To print the credit authorship contribution details
	\printcredits

	%\section*{CRediT authorship visualisation} % IMPROVED VISUAL VERSION

>>>>

@@

\usepackage[authoryear]{natbib}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
\usepackage[british]{babel}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{xurl}
\usepackage{pdfpages}
\usepackage{enumitem}
\usepackage{lineno}
\usepackage{float}
\usepackage{longtable}
\usepackage{geometry}
\usepackage{tabularx}
\usepackage{fontspec}
\setmainfont[Mapping=tex-text,Ligatures={Common,Rare,Discretionary}]{Linux Libertine}

\begin{document}
\let\WriteBookmarks\relax
\def\floatpagepagefraction{1}
\def\textpagefraction{.001}

\shorttitle{T-reX: Quantifying waste and material footprints in LCA}

\shortauthors{McDowall, S2. et~al.}

\title [mode = title]{T-reX: Quantifying waste and material footprints in current and future Life Cycle Assessment (LCA) databases}

@@

\cormark[1]
\ead{s.c.mcdowall@cml.leidenuniv.nl}
\ead[url]{scmcd.ch}
\credit{Methodology, Software, Validation, Formal analysis, Investigation, Data curation, Writing: original draft, Writing: review \& editing, Visualization}

@@

\cortext[1]{Corresponding author}

\begin{abstract}
    The quintessential principle of the \protect\TLSdel{`circular economy'} \protect\TLSins{circular economy} is to minimise material consumption and waste generation. Thus, identifying and quantifying waste and material flows is critical.

    Life Cycle Assessment (LCA) is a powerful \protect\TLSins{quantitative} method for this, \protect\TLSins{especially} given its capacity to \protect\TLSdel{pinpoint hotspots} \protect\TLSins{expose the details} of \protect\TLSdel{environmental impact throughout} an activity's \protect\TLSins{entire} life cycle, \protect\TLSdel{where} \protect\TLSins{often guiding} the implementation of circular principles \protect\TLSins{to where they} could be most effective.

\protect\TLSdel{We introduce T-reX, a Python}

    \protect\TLSins{Pursuant to these goals the authors have developed `T-reX', an analytical} tool \protect\TLSdel{extending} \protect\TLSins{for LCA, written in Python. T-reX extends on the existing computational ecosystem centered on \texttt{Brightway} and \texttt{Premise} and can facilitate} the \protect\TLSdel{brightway ecosystem, to allow facile} quantification of user-defined supply chain demands \protect\TLSins{for activities} in current and prospective scenarios. T-reX streamlines database manipulation for LCA \protect\TLSdel{practitioners and integrates} \protect\TLSins{practitioners, integrating} methods to aggregate and analyse \protect\TLSdel{waste and material flows, facilitating rapid hotspot identification.

A} \protect\TLSins{demand inventories that could expose hotspots of hidden risk.

    With a simple} case study \protect\TLSdel{of five} \protect\TLSins{using} lithium-ion \protect\TLSdel{batteries demonstrates} \protect\TLSins{batteries, we demonstrate some of} T-reX's \protect\TLSdel{utility, quantifying categorised} \protect\TLSins{potential by detailing and exploring aspects of their} waste and material \protect\TLSdel{inventory} \protect\TLSins{footprints. Since such} footprints \protect\TLSdel{and their potential environmental burdens.} \protect\TLSins{are often linked to negative externalities,} T-reX \protect\TLSdel{can aid} \protect\TLSins{could support} sustainable decision-making and contribute to the development of \protect\TLSdel{`circular economy'} \protect\TLSins{the circular economy} by \protect\TLSdel{facilitating analysis} \protect\TLSins{promoting the understanding} of \protect\TLSins{our} material consumption and waste \protect\TLSdel{generation in LCA.} \protect\TLSins{generation.}
\end{abstract}

\begin{graphicalabstract}
	\includegraphics[width=1\columnwidth]{figs/grabs_T-reX_graphical-abstract.pdf}
	\end{graphicalabstract}

\begin{highlights}
	\item T-reX, a new open-source program for waste \& material footprints in Life Cycle Assessment (LCA)
	\item Assesses supply risks by calculating demand inventories for critical materials
	\item Simplifies quantification of user-specified waste and material footprint categories
	\item Rapidly identifies hotspots of waste generation and material demand in supply	chains
	\item Presents a case study of five lithium-ion batteries to demonstrate T-reX's potential
\end{highlights}

\begin{keywords}
circular economy\sep waste footprint\sep material footprint\sep life cycle assessment\sep critical raw materials \sep supply chain
\end{keywords}

@@

\begin{table*}[!htbp]
	\centering
	\caption{List of abbreviations}\label{tab:abbreviations}
	\begin{tabular}{ll}
		\toprule
		CCS	              & Carbon Capture and Storage                                          \\
		CF                & Carbon Footprint                                                    \\
		EF                & Ecological Footprint                                                \\
		CML               & Centrum voor Milieuwetenschappen (Centre for Environmental Science) \\
		CRM               & Critical Raw Material                                               \\
		CPC               & Cooperative Patent Classification                                   \\
		CSI               & Crustal Scarcity Indicator                                          \\
		CSP               & Crustal Scarcity Potential                                          \\
		EDIP              & Environmental Design of Industrial Products                         \\
		EEIOA             & Environmentally Extended Input-Output Analysis                      \\
		EOL               & End-of-Life                                                         \\
		FOEN              & Swiss Federal Office for the Environment                            \\
		IAM               & Integrated Assessment Model                                         \\
		IE                & Industrial Ecology                                                  \\
		LCA               & Life Cycle Assessment                                               \\
		LCI               & Life Cycle Inventory                                                \\
		LCIA              & Life Cycle Impact Assessment                                        \\
		LFP               & Lithium Iron Phosphate                                              \\
		Li-ion            & Lithium-ion                                                         \\
		LiMn\(_2\)O\(_4\) & Lithium Manganese Oxide                                             \\
		MF                & Material Footprint                                                  \\
		MFA               & Material Flow Analysis                                              \\
		NCA               & Nickel Cobalt Aluminum                                              \\
		NMC               & Nickel Manganese Cobalt                                             \\
		PWF               & Product Waste Footprint                                             \\
		RCP               & Representative Concentration Pathway                                \\
		ReCiPe            & A standard LCIA method set                                          \\
		SDG               & Sustainable Development Goals                                       \\
		SSPs              & Shared Socioeconomic Pathways                                       \\
		T-reX             & The Tool for analysing re-X in LCA                                  \\
		UBP               & Umweltbelastungspünkte (Environmental Impact Points)                \\
		UN                & United Nations                                                      \\
		WF                & Waste Footprint                                                     \\
		\bottomrule
	\end{tabular}
\end{table*}

\section{Introduction}\label{sec:introduction}  
\subsection[[Introduction to T-reX]]{Introduction to T-reX}\label{sec:intro-trex}

@@

\subsection[[Background and need]]{Background and need}\label{sec:intro-background}

@@

\subsection*{}

@@

\begin{figure*}[]
	\centering
	\includegraphics[width=1.8\columnwidth]{figs/T-reX_method.pdf}
	\caption{The workflow of T-reX presented in a generalised (a) and computational (b) format. Application of T-reX to one or more LCA databases creates a \texttt{brightway} project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply \protect\TLSdel{chain.
	\label{fig:methods-flowchart}} chain.}\label{fig:methods-flowchart}
\end{figure*}

@@

\protect\TLSdel{We}
\protect\TLSins{To provide an example, we} investigated five types of Li-ion batteries in the
\texttt{ecoinvent} 3.9.1 `cutoff' LCA database~\citep{ecoinvent2016version3},
each represented by their unamended global market \protect\TLSdel{inventories:
[itemsep=0pt]} \protect\TLSins{inventories. The system
boundaries. These following markets were used for the example calculations and
unaltered by the authors:

}\begin{itemize}
	\item Li-ion, LFP, rechargeable, prismatic
	\item Li-ion, LiMn\(_2\)O\(_4\), rechargeable, prismatic
	\item Li-ion, NMC111, rechargeable, prismatic
	\item Li-ion, NMC811, rechargeable, prismatic
	\item Li-ion, NCA, rechargeable, prismatic
\end{itemize}

@@

\protect\TLSdel{[itemsep=0pt]}
\begin{itemize}
	\item ReCiPe 2016 v1.03, midpoint (I)~\citep{huijbregts2016recipe}
	\item \protect\TLSdel{EF v3.0 no LT~} \protect\TLSins{Environmental Footprint (EF) v3.1 (no long-term emissions)~}\citep{eu2023ef}
	\item \protect\TLSins{Spatial differentiation in life cycle impact assessment---the} EDIP 2003 no \protect\TLSdel{LT~} \protect\TLSins{long term emissions method~}\citep{hauschild2003edip}
	\item \protect\TLSdel{CSI} \protect\TLSins{The crustal scarcity indicator method---CSI} 2020~\citep{arvidsson2020csi}
\end{itemize}

@@

\subsubsection{LCA calculations}
For each combination of activity, method, and database, a \protect\TLSdel{single score} \protect\TLSins{total summation} was
calculated along with details of the top contributing processes. Additionally,
for the T-reX methods, a contribution analysis was \protect\TLSdel{performed. This involved utilising} \protect\TLSins{performed with} the
\texttt{bwa.compare\_activities\_by\_grouped\_leaves} function from the
\texttt{brightway2\_analyzer} package~\citep{mutel2016brightway2analyzer}, a
component of the \texttt{brightway} ecosystem. This function performs \protect\TLSins{one
subjective method of} graph traversal on the impact matrix of the LCA object to
a specified \protect\TLSdel{cutoff and} \protect\TLSins{cutoff. The} groups \protect\TLSins{of} the resulting \protect\TLSdel{leaves} \protect\TLSins{`leaves' are collected} by their
Cooperative Patent Classification (CPC) \protect\TLSdel{codes. This provides} \protect\TLSins{codes [giving a representation of
industry sector involvement]. T-reX provides, essentially, a tabulation of
internal exchanges and an} insight into the products and sectors in the supply
chain of the activity that carry the most substantial responsibility for the
final waste generation or material demand footprints.

@@

An example of the output from the application of T-reX has been included in the
supplementary \protect\TLSdel{material in~\autoref{sec:supplementary}.} \protect\TLSins{material~\autoref{sec:supplementary-5}.} The
manipulated \texttt{ecoinvent} \protect\TLSdel{databases, which} \protect\TLSins{databases (which} are the main product of \protect\TLSdel{T-reX,} \protect\TLSins{T-reX)}
can be recreated by following the instructions in the package
documentation~\citep{mcdowall2023T-reXdocs}.

@@

As described in~\autoref{sec:method-casestudy}, this case study calculated the
waste and material footprints (with a variety of other indicators) for the
unaltered background inventories of five Li-ion batteries with the functional
unit being 1~kg of the battery at the global market. The purpose of this simple
case study was to test, verify, and demonstrate the functionality and
limitations of T-reX. This section includes some highlights of the results,
with full results included in the supplementary material \protect\TLSdel{(\autoref{sec:supplementary}).}
\protect\TLSins{(\autoref{sec:supplementary-7} and \autoref{sec:supplementary-8}).} Because the `pseudo-LCIA' methods created by
T-reX are integrated into the user's \texttt{brightway} project as if they were
standard LCIA methods, the footprint calculations can be performed in the
customary way. In the supplementary material, there are screenshots of selected
results obtained using the \texttt{ActivityBrowser} software, including
contribution analysis and a Sankey diagram that \protect\TLSdel{disaggregates} \protect\TLSins{dis-aggregates} the final
footprint result over the activities in the supply chain.

\subsubsection[Sankey visualisation of  flow inventory footprints]{Sankey visualisation of \protect\TLSdel{flows in waste and material} \protect\TLSins{flow} inventory footprints}\label{sec:results-case_study-sankey}

\autoref{fig:sankey_waste} presents a \protect\TLSins{heavily abridged} Sankey diagram depicting the supply chain flows of `total solid waste' attributed to the NMC 811 Li-ion battery at the global market in the \texttt{ecoinvent} database version 3.9.1. As the supply chains are expansive, this visualisation has been greatly simplified \protect\TLSins{to be legible} for print, \protect\TLSdel{the full version is} \protect\TLSins{multiple versions are} included in the supplementary \protect\TLSdel{material~\autoref{sec:supplementary}.} \protect\TLSins{material in~\autoref{sec:supplementary-8}.} As is evident in the figure, the largest contributors to the waste footprint are the extraction and processing of the raw materials, with the largest flows coming from the extraction of cobalt, lithium, and nickel. Due to the nature of LCA, the waste flows do not represent the `real' physical flows directly, but rather an abstraction thereof based on, in this case, economic allocation---the market price of the functional unit produced by the activity~\citep{guinee2004economicallocation}. If an unallocated database were available, the waste flows could be calculated with T-reX in the same way and \protect\TLSins{would} be more directly related to the physical flows of the activities.

\protect\TLSdel{[htbp]}
\begin{figure}[!htbp]
	\centering

	\includegraphics[width=0.8\columnwidth]{figs/T-reX_NMC811_WasteTotalSolid.pdf}
	\caption{A \protect\TLSins{greatly} simplified Sankey diagram \protect\TLSins{from \texttt{ActivityBrowser}} showing the total solid waste footprint flows for the
	 NMC 811 battery supply chain at the global market in the \texttt{ecoinvent} database version \protect\TLSdel{3.9.1.}\label{fig:sankey_waste}} 3.9.1}\label{fig:sankey_waste}
\end{figure}

@@

\autoref{fig:waste_totalANDcarbon} (a) shows the `total solid waste' inventory footprint for the five Li-ion batteries in the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model. The NMC811 battery has the largest footprint, producing over 50~kg of waste per kilogram of battery produced. The  LiMn\(_2\)O\(_4\) battery has the smallest footprint, producing less than 4~kg of waste per kilogram of battery. In each case, there was only a slight downward trend in the waste footprints between 2020 and 2100. This is mostly in the period between 2020 and 2040 and could be attributable to the relatively rapid decrease in \protect\TLSdel{fossil-fuel} \protect\TLSins{fossil fuel} use that is factored into the models over this time (since many produce relatively large amounts of waste during extraction and combustion). For the total waste generated by these batteries, there was very little difference observed between the baseline and PkBudg500 RCPs.

The inclusion of carbon capture and storage (CCS) in the prospective databases
is apparent in~\autoref{fig:waste_totalANDcarbon} (b), with the rapid increase
in the production of carbon dioxide `waste' over the period from 2020--2040
that is not seen in the baseline scenario. This result highlights the fact that
the (often) downward trends in global warming impacts calculated with
prospective databases using standard LCIA methods are dependent on the
assumptions made about the introduction of CCS technology. The actual
deployment of these technologies was only around 37~Mt \protect\TLSdel{CO$_2$/yr} \protect\TLSins{CO\(_2\)/yr} as of
2023~\citep{dziejarski2023ccs}, far short of the levels projected in many of
the RCP scenarios~\citep{sacchi2023premisedocs}. The authors consider that the \protect\TLSdel{over representation}
\protect\TLSins{over-representation} of CCS and the \protect\TLSdel{under representation} \protect\TLSins{under-representation} of waste processing
technologies in the prospective databases is one significant limitation to the
validity of prospective LCA databases.

\protect\TLSdel{[htbp]} \protect\TLSins{Significant work is needed to improve
the representation of these technologies in the databases to ensure that the
results are more realistic and useful for decision-making.

}\begin{figure*}[!htbp]
	\centering

	\includegraphics[width=1.8\columnwidth]{figs/T-reX-wastefootprint-totalANDcarbon.pdf}
	\caption{T-reX calculated supply chain inventories for the five Li-ion batteries in the case study, showing their `footprints' for: (a) total solid waste, and (b) carbon dioxide waste \protect\TLSdel{(from} \protect\TLSins{(produced from} carbon capture and storage (CCS) only). The footprints were modelled with prospective LCA databases from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:waste_totalANDcarbon}
\end{figure*}

\subsubsection[Contribution of `top-processes' in the supply  chain}\label{sec:results-case_study-topprocesses}

\autoref{fig:top_contribution]{Contribution of `top-processes' in the supply \protect\TLSdel{chain} \protect\TLSins{chain}\label{sec:results-case_study-topprocesses}}

@@

\protect\TLSdel{[htbp]}
\begin{figure}[!htbp]
	\centering

	\includegraphics[width=0.8\columnwidth]{figs/T-reX-wastefootprint-processcontributions.pdf}
	\caption{Contribution by the `top-processes' to graphite demand inventory of the LiMn\(_2\)O\(_4\) battery from 2020 to 2100 under the SSP2 scenario using the baseline RCPs of the REMIND \protect\TLSdel{model.}\label{fig:top_contribution}} model}\label{fig:top_contribution}
\end{figure}

@@

\protect\TLSdel{[htbp]}
\begin{figure}
	\centering

	\includegraphics[width=0.8\columnwidth]{figs/T-reX-wastefootprint-sectorcontributions.pdf}
	\caption{Contribution of industrial sectors to the natural gas footprint of the LFP battery from 2020 to 2100 under the SSP2 scenario using the baseline RCP of the REMIND \protect\TLSdel{model.}\label{fig:cpc_contribution}} model}\label{fig:cpc_contribution}
\end{figure}

@@

A comparison of the results \protect\TLSdel{from T-reX's} \protect\TLSins{for} `Coal (black)' demand \protect\TLSins{using the T-reX} method
with the LCIA method `EDIP 2003 \protect\TLSdel{-} \protect\TLSins{–} coal no LT' is shown
in~\autoref{fig:comparison_methods}. In this case, both the trends and the
magnitude of the scores were very similar, both demonstrating a general
decrease in the use of coking coal in the battery's footprints. Such
comparability was also observed for other fossil-fuel-related \protect\TLSdel{methods} \protect\TLSins{demands} (e.g.,
natural gas and petroleum), and to a lesser extent, for some of the metal
demand methods (e.g., zinc and cobalt). Correlation between standard LCIA
methods and T-reX methods is not generally to be expected, however, due to
fundamental differences in the way that the methods are constructed. In
standard LCIA methods, impact scores are derived exclusively from the magnitude
of the exchanges between the biosphere and the technosphere, that is,
extraction and emission. In the T-reX methods, the footprint scores are based
on an accounting of either waste generated or material demand, both of which
are technosphere-technosphere exchanges in terms of LCA modelling. For the
material demand methods especially, this distinction is critical. For example,
the application of the EDIP 2003 or the CSP methods for a given metal will
provide a score that is proportional to the amount extracted by mining, whereas
the T-reX method provides an aggregation of the exchanges with the market for
that metal. The T-reX method, therefore, considers cases of co-production,
recycling, and substitution, providing a picture of the supply chain pressures
that are not captured by the standard LCIA methods. This makes the T-reX
methods more sensitive to the modeling choices \protect\TLSdel{(e.g.,economic} \protect\TLSins{(e.g., economic} allocation) that
are generally embedded in LCA databases.

\protect\TLSdel{[htbp]}
\begin{figure*}[!htbp]
	\centering

	\includegraphics[width=1.8\columnwidth]{figs/T-reX-coalANDedip.pdf}
	\caption{Comparison of the (a) `T-reX \protect\TLSdel{-} \protect\TLSins{–} Coal (black)' demand footprint with the (b) conventional LCIA method `EDIP 2003 \protect\TLSdel{-} \protect\TLSins{–} coal no \protect\TLSdel{LT'  in} \protect\TLSins{longterm'. In} the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model for all five batteries.}\label{fig:comparison_methods}
\end{figure*}

@@

\printcredits
